---
title: "Component Developmental Sequence Analysis"
author: "Bryan Terlizzi"
output: html_notebook
---

```{r include=FALSE}
library(tidyverse)
library(here)
library(dplyr)
library(stringr)
library(purrr)

ds <- read.csv(here("Data", "Merged_CompSkillSeq_byTimepoint_v2.csv"), na.strings = c("", "NA"))

```


# Data Description:

Intervention and Control Cohorts across 5 timepoints. Component Developmental sequence data across 4 skills (Standing Long Jump, 1-Leg Hop, Throwing, and Kicking). Collected at 5 different timepoints.

### Created Variables across the 4 skills:
1) Mode for each DS component. 
2) Component Profile for each of 5 trial.
3) Component Profile Mode for each of 5 timepoints.


```{r echo=FALSE}
#Creating Mode for each component

# Define mode function
get_mode <- function(x) {
  x <- na.omit(x)
  if (length(x) == 0) return(NA)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Compute modal component scores
ds <- ds %>%
  rowwise() %>%
  mutate(
    # Throwing
    Throw_Step_Mode = get_mode(c_across(starts_with("Throw_Step_"))),
    Throw_Trunk_Mode = get_mode(c_across(starts_with("Throw_Trunk_"))),
    Throw_Humerus_Mode = get_mode(c_across(starts_with("Throw_Humerus_"))),
    Throw_Forearm_Mode = get_mode(c_across(starts_with("Throw_Forearm_"))),

    # Kicking
    Kick_Approach_Mode = get_mode(c_across(starts_with("Kick_Approach_"))),
    Kick_Support.Loc_Mode = get_mode(c_across(starts_with("Kick_Support.Loc_"))),
    Kick_Knee.Action_Mode = get_mode(c_across(starts_with("Kick_Knee.Action_"))),
    Kick_Trunk.Action_Mode = get_mode(c_across(starts_with("Kick_Trunk.Action_"))),
    Kick_Arm.Action_Mode = get_mode(c_across(starts_with("Kick_Arm.Action_"))),
    Kick_Follow.Through_Mode = get_mode(c_across(starts_with("Kick_Follow.Through_"))),

    # Hopping
    Hopping_Arm_Mode = get_mode(c_across(starts_with("Hopping_Arm_"))),
    Hopping_Leg_Mode = get_mode(c_across(starts_with("Hopping_Leg_"))),

    # Jumping
    Jump_TO_Arm_Mode = get_mode(c_across(starts_with("Jump_TO_Arm_"))),
    Jump_TO_Leg_Mode = get_mode(c_across(starts_with("Jump_TO_Leg_"))),
    Jump_LND_Arm_Mode = get_mode(c_across(starts_with("Jump_LND_Arm_"))),
    Jump_LND_Foot_Mode = get_mode(c_across(starts_with("Jump_LND_Foot_"))),
    Jump_LND_Shank_Mode = get_mode(c_across(starts_with("Jump_LND_Shank_")))
  ) %>%
  ungroup()


# Creating Dev Seq Profile variables for each trial

ds <- ds %>%
  mutate(
    Throw_Profile_1 = str_c(Throw_Step_1, Throw_Trunk_1, Throw_Humerus_1, Throw_Forearm_1, sep = ""),
    Throw_Profile_2 = str_c(Throw_Step_2, Throw_Trunk_2, Throw_Humerus_2, Throw_Forearm_2, sep = ""),
    Throw_Profile_3 = str_c(Throw_Step_3, Throw_Trunk_3, Throw_Humerus_3, Throw_Forearm_3, sep = ""),
    Throw_Profile_4 = str_c(Throw_Step_4, Throw_Trunk_4, Throw_Humerus_4, Throw_Forearm_4, sep = ""),
    Throw_Profile_5 = str_c(Throw_Step_5, Throw_Trunk_5, Throw_Humerus_5, Throw_Forearm_5, sep = ""),
    Hopping_Profile_1 = str_c(Hopping_Arm_1, Hopping_Leg_1, sep = ""),
    Hopping_Profile_2 = str_c(Hopping_Arm_2, Hopping_Leg_2, sep = ""),
    Hopping_Profile_3 = str_c(Hopping_Arm_3, Hopping_Leg_3, sep = ""),
    Hopping_Profile_4 = str_c(Hopping_Arm_4, Hopping_Leg_4, sep = ""),
    Kick_Profile_1 = str_c(Kick_Approach_1, Kick_Support.Loc_1, Kick_Knee.Action_1, Kick_Trunk.Action_1, Kick_Arm.Action_1, Kick_Follow.Through_1, sep = ""),
    Kick_Profile_2 = str_c(Kick_Approach_2, Kick_Support.Loc_2, Kick_Knee.Action_2, Kick_Trunk.Action_2, Kick_Arm.Action_2, Kick_Follow.Through_2, sep = ""),
    Kick_Profile_3 = str_c(Kick_Approach_3, Kick_Support.Loc_3, Kick_Knee.Action_3, Kick_Trunk.Action_3, Kick_Arm.Action_3, Kick_Follow.Through_3, sep = ""),
    Kick_Profile_4 = str_c(Kick_Approach_4, Kick_Support.Loc_4, Kick_Knee.Action_4, Kick_Trunk.Action_4, Kick_Arm.Action_4, Kick_Follow.Through_4, sep = ""),
    Kick_Profile_5 = str_c(Kick_Approach_5, Kick_Support.Loc_5, Kick_Knee.Action_5, Kick_Trunk.Action_5, Kick_Arm.Action_5, Kick_Follow.Through_5, sep = ""),
    Jump_TO_Profile_1 = str_c(Jump_TO_Arm_1, Jump_TO_Leg_1, sep = ""),
    Jump_TO_Profile_2 = str_c(Jump_TO_Arm_2, Jump_TO_Leg_2, sep = ""),
    Jump_TO_Profile_3 = str_c(Jump_TO_Arm_3, Jump_TO_Leg_3, sep = ""),
    Jump_TO_Profile_4 = str_c(Jump_TO_Arm_4, Jump_TO_Leg_4, sep = ""),
    Jump_TO_Profile_5 = str_c(Jump_TO_Arm_5, Jump_TO_Leg_5, sep = ""),
    Jump_LND_Profile_1 = str_c(Jump_LND_Arm_1, Jump_LND_Foot_1, Jump_LND_Shank_1, sep = ""),
    Jump_LND_Profile_2 = str_c(Jump_LND_Arm_2, Jump_LND_Foot_2, Jump_LND_Shank_2, sep = ""),
    Jump_LND_Profile_3 = str_c(Jump_LND_Arm_3, Jump_LND_Foot_3, Jump_LND_Shank_3, sep = ""),
    Jump_LND_Profile_4 = str_c(Jump_LND_Arm_4, Jump_LND_Foot_4, Jump_LND_Shank_4, sep = ""),
    Jump_LND_Profile_5 = str_c(Jump_LND_Arm_5, Jump_LND_Foot_5, Jump_LND_Shank_5, sep = "")
  )
    

#Combining Takeoff and Landing for a total Jump Profile (too variable)

ds <- ds %>%
 mutate(
  Jump_Profile_1 = str_c(Jump_TO_Profile_1, Jump_LND_Profile_1),
  Jump_Profile_2 = str_c(Jump_TO_Profile_2, Jump_LND_Profile_2),
  Jump_Profile_3 = str_c(Jump_TO_Profile_3, Jump_LND_Profile_3),
  Jump_Profile_4 = str_c(Jump_TO_Profile_4, Jump_LND_Profile_4),
  Jump_Profile_5 = str_c(Jump_TO_Profile_5, Jump_LND_Profile_5)
  )



#Creating Modal Profiles for each skill

ds <- ds %>%
  rowwise() %>%
  mutate(
    Throw_Profile_Mode = get_mode(c_across(starts_with("Throw_Profile_"))),
    Hopping_Profile_Mode = get_mode(c_across(starts_with("Hopping_Profile_"))),
    Kick_Profile_Mode = get_mode(c_across(starts_with("Kick_Profile_"))),
    Jump_Profile_Mode = get_mode(c_across(starts_with("Jump_Profile_"))),
    Jump_TO_Profile_Mode = get_mode(c_across(starts_with("Jump_TO_Profile_"))),
    Jump_LND_Profile_Mode = get_mode(c_across(starts_with("Jump_LND_Profile_")))
  ) %>%
  ungroup()


```

# Analysis. 

## Frequency of Each Component for each timepoint 

```{r echo=FALSE}
#Table function for putting together the tables

component_freq_table <- function(data, component_mode_col) {
  data %>%
    filter(!is.na(.data[[component_mode_col]])) %>%
    group_by(Cohort, Timepoint, Score = as.character(.data[[component_mode_col]])) %>%
    summarise(Frequency = n(), .groups = "drop") %>%
    group_by(Cohort, Timepoint) %>%
    mutate(Proportion = Frequency / sum(Frequency)) %>%
    ungroup() %>%
    mutate(Proportion = as.numeric(Proportion)) %>%
    select(Cohort, Timepoint, Score, Proportion) %>%
    pivot_wider(names_from = Score, values_from = Proportion, values_fill = 0) %>%
    arrange(Cohort, Timepoint)
}

# ---- BARPLOTS TO SHOW PROPORTION OF COMPONENT SCORES ----

# 1. List of component mode variable names
component_vars <- c(
  "Throw_Step_Mode", "Throw_Trunk_Mode", "Throw_Humerus_Mode", "Throw_Forearm_Mode",
  "Kick_Approach_Mode", "Kick_Support.Loc_Mode", "Kick_Knee.Action_Mode", "Kick_Trunk.Action_Mode",
  "Kick_Arm.Action_Mode", "Kick_Follow.Through_Mode",
  "Hopping_Arm_Mode", "Hopping_Leg_Mode",
  "Jump_TO_Arm_Mode", "Jump_TO_Leg_Mode",
  "Jump_LND_Arm_Mode", "Jump_LND_Foot_Mode", "Jump_LND_Shank_Mode"
)

# 2. Initialize list to store plots
component_barplots <- list()

# 3. Loop through each component
for (component in component_vars) {
  
  # Wide to long table for plotting
  long_df <- component_freq_table(ds, component) %>%
    pivot_longer(cols = -c(Cohort, Timepoint), names_to = "Score", values_to = "Proportion") %>%
    mutate(Score = factor(Score), Timepoint = factor(Timepoint))
  
  # Plot
  p <- ggplot(long_df, aes(x = Timepoint, y = Proportion * 100, fill = Score)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    facet_wrap(~ Cohort, ncol = 1, labeller = labeller(Cohort = c(C1 = "Cohort 1", C2 = "Cohort 2"))) +
    scale_y_continuous(name = "Proportion of Participants (%)", limits = c(0, 100)) +
    labs(
      title = paste("Component Score Proportions:", component),
      x = "Timepoint",
      fill = "Component Score"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      strip.text = element_text(face = "bold", size = 13)
    )
  
  component_barplots[[component]] <- p
}

# Example: Show plot
#component_barplots$Throw_Step_Mode

```

### Throwing. 

Step. 

```{r echo=FALSE}
#Throwing
component_freq_table(ds, "Throw_Step_Mode")
component_barplots$Throw_Step_Mode

```

Trunk.   

```{r echo=FALSE}
component_freq_table(ds, "Throw_Trunk_Mode")
component_barplots$Throw_Trunk_Mode
```

Humerus.  

```{r echo=FALSE}
component_freq_table(ds, "Throw_Humerus_Mode")
component_barplots$Throw_Humerus_Mode
```

Forearm.  

```{r echo=FALSE}
component_freq_table(ds, "Throw_Forearm_Mode")
component_barplots$Throw_Forearm_Mode
```

### Kicking

Approach.  
```{r echo=FALSE}
#Kicking
component_freq_table(ds, "Kick_Approach_Mode")
component_barplots$Kick_Approach_Mode
```

Support Location.  

```{r echo=FALSE}
component_freq_table(ds, "Kick_Support.Loc_Mode")
component_barplots$Kick_Support.Loc_Mode
```

Knee Action.  

```{r echo=FALSE}
component_freq_table(ds, "Kick_Knee.Action_Mode")
component_barplots$Kick_Knee.Action_Mode
```

Trunk Action.  

```{r echo=FALSE}
component_freq_table(ds, "Kick_Trunk.Action_Mode")
component_barplots$Kick_Trunk.Action_Mode
```

Arm Action.  

```{r echo=FALSE}
component_freq_table(ds, "Kick_Arm.Action_Mode")
component_barplots$Kick_Arm.Action_Mode
```

Follow Through.  

```{r echo=FALSE}
component_freq_table(ds, "Kick_Follow.Through_Mode")
component_barplots$Kick_Follow.Through_Mode
```

### Hopping.  

Arm Action.  
```{r echo=FALSE}
#Hopping
component_freq_table(ds, "Hopping_Arm_Mode")
component_barplots$Hopping_Arm_Mode
```

Leg Action.  

```{r echo=FALSE}
component_freq_table(ds, "Hopping_Leg_Mode")
component_barplots$Hopping_Leg_Mode
```

### Jumping.  

Takeoff - Arm.  
```{r echo=FALSE}
#Jumping
component_freq_table(ds, "Jump_TO_Arm_Mode")
component_barplots$Jump_TO_Arm_Mode
```

Takeoff - Leg.  

```{r echo=FALSE}
component_freq_table(ds, "Jump_TO_Leg_Mode")
component_barplots$Jump_TO_Leg_Mode
```

Landing - Arm.  

```{r echo=FALSE}
component_freq_table(ds, "Jump_LND_Arm_Mode")
component_barplots$Jump_LND_Arm_Mode
```

Landing - Foot.  

```{r echo=FALSE}
component_freq_table(ds, "Jump_LND_Foot_Mode")
component_barplots$Jump_LND_Foot_Mode
```

Landing - Shank.  

```{r echo=FALSE}
component_freq_table(ds, "Jump_LND_Shank_Mode")
component_barplots$Jump_LND_Shank_Mode
```



## Frequency and Proportion of Profile Modes at different timepoints

```{r echo=FALSE}
#Creating function that puts together modes for different profiles

modal_freq_table1 <- function(data, profile_col) {
  data %>%
    filter(!is.na(.data[[profile_col]])) %>%  # exclude NA modal profiles
    count(Cohort, Timepoint, !!sym(profile_col), name = "Frequency") %>%
    group_by(Cohort, Timepoint) %>%
    mutate(Proportion = Frequency / sum(Frequency)) %>%
    arrange(Cohort, Timepoint, desc(Frequency)) %>%
    ungroup()
}

modal_freq_table <- function(data, profile_col) {
  # Count + proportion
  freq_data <- data %>%
    filter(!is.na(.data[[profile_col]])) %>%
    count(Cohort, Timepoint, Profile = .data[[profile_col]], name = "Frequency") %>%
    group_by(Cohort, Timepoint) %>%
    mutate(Proportion = Frequency / sum(Frequency)) %>%
    arrange(Cohort, Timepoint, desc(Proportion)) %>%
    slice_head(n = 5) %>%  # keep top 5 per Cohort × Timepoint
    ungroup()

  # Pivot to wide
  freq_data %>%
    select(Cohort, Timepoint, Profile, Proportion) %>%
    pivot_wider(
      names_from = Profile,
      values_from = Proportion,
      values_fill = 0
    ) %>%
    arrange(Cohort, Timepoint)
}


```


Try at a Figure to illustrate

```{r echo=FALSE}
# Function: Get top 5 modal profiles per timepoint (long format)
modal_freq_long <- function(data, profile_col) {
  data %>%
    filter(!is.na(.data[[profile_col]])) %>%
    count(Cohort, Timepoint, Profile = .data[[profile_col]], name = "Frequency") %>%
    group_by(Cohort, Timepoint) %>%
    mutate(Proportion = Frequency / sum(Frequency)) %>%
    arrange(Cohort, Timepoint, desc(Proportion)) %>%
    slice_head(n = 5) %>%
    ungroup()
}

# List of skill profile variables
skill_profiles <- c(
  "Throw_Profile_Mode",
  "Hopping_Profile_Mode",
  "Kick_Profile_Mode",
  "Jump_Profile_Mode",
  "Jump_TO_Profile_Mode",
  "Jump_LND_Profile_Mode"
)

# Initialize empty list to store plots
skill_profile_plots <- list()

# Loop through each skill, create plot, and save in list
for (profile in skill_profiles) {
  
  # Get long format data
  plot_data <- modal_freq_long(ds, profile)
  
  # Create plot
  p <- ggplot(plot_data, aes(x = Timepoint, y = Proportion, color = Profile, group = Profile)) +
    geom_line(size = 1.1) +
    geom_point(size = 2.5) +
    facet_wrap(~ Cohort, ncol = 1, labeller = labeller(Cohort = c(C1 = "Cohort 1 (C1)", C2 = "Cohort 2 (C2)"))) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(
      title = paste("Top 5 Modal Profiles Over Time:", str_replace_all(profile, "_", " ")),
      x = "Timepoint",
      y = "Proportion of Participants",
      color = "Profile"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      strip.text = element_text(face = "bold", size = 14, color = "black"),
      plot.title = element_text(face = "bold", hjust = 0.5),
      panel.spacing.y = unit(1.5, "lines")
    )
  # Store plot
  skill_profile_plots[[profile]] <- p
}

# Example: display one
skill_profile_plots$Throw_Profile_Mode
```
**Note:** Throwing profile components are ordered as: Step, Trunk, Humerus, Forearm. 

```{r echo=FALSE}
modal_freq_table(ds, "Throw_Profile_Mode")
```
**Note:** Throwing profile components are ordered as: Step, Trunk, Humerus, Forearm. 

```{r echo=FALSE}
skill_profile_plots$Kick_Profile_Mode
```
**Note:** Kicking profile components are ordered as: Approach, Support Leg, Knee Action, Trunk Action, Arm Action, Follow Through. 

```{r echo=FALSE}
modal_freq_table(ds, "Kick_Profile_Mode")
```
**Note:** Kicking profile components are ordered as: Approach, Support Leg, Knee Action, Trunk Action, Arm Action, Follow Through. 

```{r echo=FALSE}
skill_profile_plots$Hopping_Profile_Mode
```
**Note:** Hopping profile components are ordered as: Arm, Leg. 

```{r echo=FALSE}
modal_freq_table(ds, "Hopping_Profile_Mode")
```
**Note:** Hopping profile components are ordered as: Arm, Leg. 


```{r echo=FALSE}
skill_profile_plots$Jump_Profile_Mode
```
**Note:** Jump profile components are ordered as: Takeoff (Arm, Leg) + Landing (Arm, Foot, Shank). 

```{r echo=FALSE}
modal_freq_table(ds, "Jump_Profile_Mode")
```
**Note:** Jump profile components are ordered as: Takeoff (Arm, Leg) + Landing (Arm, Foot, Shank). 


```{r echo=FALSE}
skill_profile_plots$Jump_TO_Profile_Mode
```
**Note:** Jump profile components are ordered as: Takeoff (Arm, Leg) + Landing (Arm, Foot, Shank). 


```{r echo=FALSE}
modal_freq_table(ds, "Jump_TO_Profile_Mode")

```
**Note:** Jump profile components are ordered as: Takeoff (Arm, Leg) + Landing (Arm, Foot, Shank). 


```{r echo=FALSE}
skill_profile_plots$Jump_LND_Profile_Mode
```

**Note:** Jump profile components are ordered as: Takeoff (Arm, Leg) + Landing (Arm, Foot, Shank). 




```{r echo=FALSE}
modal_freq_table(ds, "Jump_LND_Profile_Mode")
```
**Note:** Jump profile components are ordered as: Takeoff (Arm, Leg) + Landing (Arm, Foot, Shank). 



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

